### 2013-07-24 doc/dissertation/document/chapters/conclusion/conclusion.tex
adding most of the future work section

@@

Standalone Monte Carlo as a preconditioner? (e.g. for flex-GMRES)

Stochastic variants of conventional preconditioners?

Variance reduction for preconditioning?

\subsection{Breaking away from $\rho(H) < 1$}
\label{subsec:future_spec_rad}

Monte Carlo methods of the 2nd Degree?

Stochastic projection methods?

Implementation optimization - can we approach Krylov performance?

Reduction in memory footprint

Multiple sets parallel overhead reduction

Controling MCSA histories as a function of nonlinear iteration/forcing
term?

>>>>

Throughout this work, three salient issues were observed over all
others. First, the performance of the MCSA implementation developed
for this work was severely lacking in terms of time to solution when
compared to production Krylov methods. In all cases, this difference
was several orders of magnitude. Second, the preconditioning strategy
developed by this work, although often successful in achieving
convergence for difficult problems, creates many obstacles to a
production implementation. These obstacles all primarily stem from the
dense composite operators that for the probabilities and weights for
the Monte Carlo game. Third, perhaps the most restrictive and
unattractive piece of MCSA as a whole is the simple fact that the
spectral radius of the physics operator iteration matrix be less than
one. This was demonstrated to prohibit convergence for cases where the
preconditioning was considered agressive and Krylov solvers exhibited
excellent convergence properties. We will address each one of these
problems in turn in this section and provide suggestions for their
solution.

When issues arising from preconditioning are not considered, there are
two primary issues that can be resolved in the MCSA implementation
developed by this work that will likely result in performance gains
for serial computations. First, all random walks in the implementation
have a state which changes as the random walk proceeds. These states
are those given by Eq~(\ref{eq:mc_walk_permutation}). For random walks
that are traversing a domain which has been decomposed in a parallel
environment, there are two sets of these states: one that represents
the states over the entire global domain and one that represents them
only over the local subdomain. Each state in the system gets a unique
index in both of these sets that act as absolute coordinates in the
random walk and permit the construction of the Monte Carlo estimates
and other data needed for the solution.

A fundamental flaw of the implementation developed for this work is
that those random walk states given by
Eq~(\ref{eq:mc_walk_permutation}) were formulated such that they were
always represented using the indices from the global set. However, all
data in the system (e.g. weights, probabilities, tallies, etc.) are
stored such that they are accessed using the local indexing
scheme. Because of this, each time a random walk wants to sample its
current probability distribution function, get a weight for its
transition, or add its estimate to a tally, a call must first be made
to translate the current global state of the random walk into its
local state\footnote{This is a lookup in a hash table.}. What this
then means is that we are performing orders of magnitude more
operations during the Monte Carlo sequence than if the random walk
sequence operated in the local index space rather than the global
index space. Preliminary profiling of the implementation has shown
that indeed a vast majority of the time in the Monte Carlo sequence is
spent doing this index translation. Therefore, future work should
perform the significant refactoring of the implementation required to
achieve this.

A second mode of optimization to improve performance regards how
discrete probability distribution functions are sampled in the
implementation. At each transition in the random walk sequence, the
discrete probability distribution function generated by the current
row of the physics operator in which a random walk resides is sampled
to get the next state in the sequence. This operation is also used
when the right hand side of the system is sampled to acquire the
starting state for each random walk in the problem through
Eq~(\ref{eq:adjoint_source_probability}). The current implementation
uses the inverse transform method for sampling where the a random
number is generated and the cumulative distribution function is search
to find the next discrete state. This algorithm has $O(log(N))$ time
complexity for sampling and $O(N)$ time complexity for
construction. The alias sampling method is suggested instead as it
produces $O(1)$ complexity for sampling and $O(N)$ time complexity for
construction. Profiling of the implementation shows that this
improvement could also provide quality reductions in run time.

For the parallel MCSA algorithm, overall parallel scaling results
indicated that the implementation was of good quality in both the
strong and weak scaling cases. However, for the multiple set cases in
the weak scaling analysis, it was noted that there was a growing
parallel overhead as a function of core count. This was counter
intuitive in that the size of the parallel group over which the set
reductions were occuring was fixed at all core counts and therefore we
do not expect the addition of such an operation to increase parallel
run times. It did, however, and therefore future research should
explore why the implementation for multiple sets generated parallel
overhead that increased with core count instead of remaining constant.

Nonlinear performance may also be improved by a closer relationship
between MCSA and the Newton iteration. We saw in our reseaerch how the
forcing term selected very tightly controls the convergence of the
linear model in FANM. For MCSA, in addition to the residual tolerance
that is dictated by this forcing term, convergence may also be
controlled by adjusting the number of stochastic histories computed at
each MCSA iteration. As the nonlinear iteration evolves, we may find
it useful to adjust the number of histories used at each MCSA
iteration in the same way as we adjust the convergence tolerance of
the linear model through the forcing term. Future research is
suggested to analyze the effects of the number of stochastic histories
on the convergence of the linear and nonlinear models and how this may
be controlled by the forcing term or another parameter.

With these implementation improvements perhaps MCSA and FANM timing
may approach that of production Krylov methods for equally conditioned
problems. Next we discuss potential research in the area of
preconditioning which may enable performance research for MCSA in the
area of memory consumption.

@@

In general, preconditioning developments in this work enabled much of
the desired research to be performed but also prevented some of it. Of
primary concern is the large memory footprint induced by the explicit
algebraic preconditioning strategy outlined in
\S~\ref{subsubsec:general_mcsa_preconditioning}. By preconditioning
in this way, we were able to leverage all modern algebraic
preconditioning strategies used in current physics practice. The
downside is that in order to generate weights and probabilities with
which to play the Monte Carlo game, me must first invert each of the
preconditioners which in many cases means applying their inverse over
each row in the system, and then we must perform matrix-matrix
multiply operations to build the composite physics operator.

These operations were observed to substantially increase runtimes of
MCSA and consume unacceptable amounts of memory, mostly due to the
fact that inverting the preconditioners create dense
matrices. Although a good fraction of the elements in each row of the
composite operator could be eliminated through the reduced domain
approximation developed by this work, memory costs still prevented
higher fidelity simulations from being performed. In addition, the
reduced domain approximation does no prevent the large run times
incurred for inverting the preconditioners. We therefore propose the
following three developments:

\begin{itemize}
\item \textbf{Research and develop alternative sampling schemes for
  the general left/right preconditioned Monte Carlo method not relying
  on composite matrix generation.} Developments that break away from
  Matrix-Matrix multiplications within the preconditioned solver
  framework improve scalability. This means avoiding the generation of
  the composite operator and instead playing the Monte Carlo game with
  weights and probabilities formed by considering the original linear
  operator and the preconditioners individually. This individual
  consideration may come from an alternative algebraic formulation of
  the general preconditioned MCSA method in
  Eq~(\ref{eq:left_right_mcsa}) or from an alternative random sampling
  strategy for the Monte Carlo game.
\item \textbf{Develop Monte Carlo variance reduction based
  preconditioning strategies that do not rely on an algebraic
  preconditioner.}  Constructing the complete inverse of the
  preconditioner is the core time and scalability constraint when
  modern algebraic preconditioners are used. Building the inverse for
  preconditioners where a sparsity pattern is not enforced typically
  results in a dense matrix that destroys scalability for domain
  decomposed parallel computations. Alleviating both of these issues
  by stochastic preconditioning methods will be studied. Decades of
  research and development in Monte Carlo variance reduction for
  particle transport \cite{booth_1994} may be leveraged to form
  similar strategies for linear systems. In the same way that this
  work has been able to adapt modern particle transport strategies for
  parallelism and apply them to linear systems, the same strategy may
  be applied to variance reduction techniques to improve MCSA
  convergence by modifying the weights and probabilities of the Monte
  Carlo game as an effective means of preconditioning.
\item \textbf{Study physics or PDE-based preconditioners for linear
  systems that produce a reduced-order model of the original system to
  which the Monte Carlo method may be applied.} As the Monte Carlo
  method serves to construct the solution correction vector in the
  MCSA method, the Monte Carlo method may be applied to a
  reduced-order physics or PDE-based system that captures the essence
  of the original system and still allows for acceleration of the
  iterative procedure. By reducing the order of the system, we are
  effectively generating a Monte Carlo problem within MCSA in which
  the linear operator is either more sparse, has a better-conditioned
  eigenvalue spectrum, or both. The reduced domain approximation has
  shown that a purely algebraic strategy for reducing the order of the
  Monte Carlo problem can result in improved overall timing
  performance due to increased sparsity at the cost of iterative
  performance as a result of loss of information. A more well informed
  strategy for order reduction via either an intelligent
  simplification of the PDEs that describe the system or a
  simplification of the physics involved may prove to be more
  effective strategy.
\end{itemize}

\subsection{Breaking Away from $\rho(H) < 1$}
\label{subsec:future_spec_rad}

All methods required quality preconditioning. Even a Krylov method may
be effectively useful without it. However, these methods typically
converge eventually whereas quality preconditioning may not be enough
to gaurantee convergence of MCSA. We saw in our work that in many
cases very agressive preconditioning was not enough to condition the
eigenvalues of the system into a regime where MCSA would apply. The
limitation that the spectral radius of the physics operator iteration
matrix must be less than one is by far the most prohibitive component
of MCSA. The fact that the Monte Carlo sequence is bsed on the
Richardson iteration means that without sufficient preconditioning,
even if this condition is achieved convergence may be incredibly
slow. Therefore, we propose two elements of future work that aim to
move away from this restriction, both fundamentally based on sampling
a sequence not based on the Neumann series.

\subsection{Monte Carlo Methods of the Second Degree}
\label{subsubsec:2_degree_mc}

\subsection{Stochastic Projection Methods}
\label{subsubsec:stochastic_projection_methods}
